Financialization of Commodities: an Asset Pricing Perspective

### 2024-06-11 others/paper/elsevier article/main.tex
Track 'others' folder

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[letterpaper, left = 1in, right = 1in, top = 1.5in, bottom = 1.5in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Financialization of Commodities: an Asset Pricing Perspective},
            pdfkeywords={Commodity futures; financialization; hedging pressure; financial crisis},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{jf}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

  \title{Financialization of Commodities: an Asset Pricing Perspective}
    % \author{Devraj Basu \\ Olivier Bauthéac}
  \author{ Devraj Basu  \\  Olivier Bauthéac  
	  \thanks{ Basu: University of Strathclyde, Glasgow, UK, \href{mailto:devraj.basu@strath.ac.uk}{\nolinkurl{devraj.basu@strath.ac.uk}}.  Bauthéac: University of Strathclyde, Glasgow, UK, \href{mailto:olivier.bautheac@strath.ac.uk}{\nolinkurl{olivier.bautheac@strath.ac.uk}}. 	  
	   \newline We thank seminar participants at IIM Ahmedabad, IIM Calcutta, SKEMA
Business School and University of Gottingen, as well as Griffith
inaugural alternative investment conference attendants for their
feedback. We also thank Joelle Miffre, Martijn Boons, John Fan, Robert
Bianchi, Chardin Wese Simen, Andrew Kaleel and particularly Olaf Korn
for their extremely useful comments and suggestions.  }
  }
    \date{}

\usepackage{longtable, tabu, pdflscape, booktabs, caption, setspace}
\doublespacing
\renewcommand*{\arraystretch}{0.65}

\begin{document}
\maketitle
\begin{abstract}
The last two decades have seen major changes in the nature of the
commodity markets with the influx of financial investor inflows followed
by the financial crisis. We study the effect of these events on
commodity futures pricing dynamics using commodity asset pricing
techniques. We find that a non-price based factor is best able to detect
the effects of commodity financialization, while the effects of the
financial crisis and after are most clearly visible through the
performance of the market factor. The effects of financialization seem
to go beyond mechanical effects induced by the rise of indexing. 
\newline
JEL: G11, G23, G13, Q02
\newline
Keywords: Commodity futures, financialization, hedging pressure, financial crisis

\end{abstract}

@@

\begin{itemize}
\tightlist
\item
  Devraj Basu: I am a co-author of this paper and wish to state that I
  have no conflict of interest, financial or otherwise. I do not hold
  paid or unpaid positions at any organization whose goals relate to the
  article.\\
  \includegraphics{pics/signature - Devraj.png}\\
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Olivier Bauthéac: I am a co-author of this paper and wish to state
  that I have no conflict of interest, financial or otherwise. I do not
  hold paid or unpaid positions at any organization whose goals relate
  to the article.\\
  \includegraphics{pics/signature - Olivier.png}\\
\end{itemize}

@@

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialization\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  \citep{buyuksahin_speculators_2014, singleton_investor_2013},
  institutional investors sent forth an unprecedented flow of capital
  into commodity futures markets. This phenomenon is commonly referred
  to as the financialization of commodities
  \citep{domanski_financial_2007}.} as well as the period of the 2008
financial crisis and its aftermath. Financialization appears to have
affected a number of economic sectors, including agriculture and energy,
and its impact has been hotly debated in the policy, legislative,
regulatory, and academic spheres. A vigorous legislative debate has led
to regulatory changes accompanied by an equally intense policy and
academic debate across a number of disciplines about the perceived
effects on commodity prices of the unprecedented inflow of institutional
funds brought into commodity futures markets by new index type
investors\footnote{See section \ref{background} for a detailled
  discussion.}. In response to immediate policy concerns, the academic
debate was initially framed rather narrowly. In this paper we take a
fundamentally different empirical approach that is broader in scope. We
consider the entire cross-section of actively traded commodities on
futures markets and use a futures based asset pricing framework that
includes factors constructed using both price and non-price attributes.

@@

The first factor model we consider (CHP) is based on the well-known
hedging pressure based theory
\citep{anderson_hedger_1983, chang_returns_1985, cootner_returns_1960, dusak_futures_1973, hicks_value_1939, hirshleifer_risk_1988, hirshleifer_determinants_1989, hirshleifer_hedging_1990, kolb_is_1992, keynes_treatise_1930}
which postulates that futures prices for a given commodity are inversely
related to the extent that commercial hedgers are short or long and we
use a factor mimicking portfolio that captures the impact of hedging
pressure as a systemic factor \citep{basu_capturing_2013}. Our second
factor model is based on open interest growth following
\citet{hong_what_2012} who find that open interest growth has predictive
power for individual commodity returns and again we construct a
long-short factor mimicking portfolio. Our third model is based on term
structure following \citet{szymanowska_anatomy_2014} and
\citet{fuertes_commodity_2015} who demonstrate that the term structure
factor has explanatory power for the cross-section of commodity returns.
We finally consider an equally weighted portfolio of commodities as a
proxy for the market factor. Our two price based and two non-price based
factors allow us to compare and contrast different channels that could
have led to changes in commodity price dynamics over the last twenty
years.

We study the performance of these models on a set of twenty-four US
traded commodities and six UK traded (LME exchange) base metals over the
1997/2017 period that we divide into four distinct sub-periods. The
first, the pre-financialization phase, spans the 1997/2003 period; the
second, the financialization phase, spans the 2004/2007 period\footnote{Starting
  point based on earlier studies
  \citep{baker_financialization_2014, christoffersen_factor_2014}.}; the
third, 2008/2013, encompasses both the financial crisis and the loose
monetary policy regimes that followed; and the fourth spans the period
starting in 2014 to present.\\
Consistent with earlier studies, we observe that mean returns for the US
traded commodities are sharply higher over the financialization period
compared to pre-financialization and this pattern extends to UK traded
metals.

The cross-sectional nature of financialization
\citep{basak_model_2016, cheng_financialization_2014} leads us to extend
the Keynesian individual hedging pressure paradigm by incorporating
aggregate CHP (\textbf{CHP}), a market wide measure of hedging pressure
first considered in \citet{hong_what_2012}. This paradigm implies that
returns for individual commodities should be high in periods where
\textbf{CHP} is low (\textbf{backwardation}) and low in periods where
\textbf{CHP} is high (\textbf{contango}). Over the first period the
cross sectional average returns during \textbf{backwardation} for both
the US and UK commodities are positive with those for the US
statistically significant, and significantly higher than during
\textbf{contango}, providing broad support for an aggregate version of
the Keynesian hedging pressure hypothesis. Over the second period this
pattern reverses, with both the US and UK commodities having significant
positive returns during \textbf{contango} with these returns being
significantly higher than during \textbf{backwardation}. This change in
pattern suggests that the onset of financialization altered the
traditional risk shifting nature of hedger's behaviour, which is implied
by the Keynesian hypothesis. The arrival of financial investors raises
the question of whether they functioned as liquidity providers or
liquidity demanders. Arguments have been made for both of these
possibilities with \citet{moskowitz_time_2012} arguing that financial
investors provide liquidity while \citet{Kangtaletwopremiums2017} argue
that they demand liquidity. Our analysis of returns in periods of low
and high aggregate CHP before and during financialization allows us to
analyse this issue from a new perspective and suggests that financial
investors may have been liquidity demanders leading hedgers to become
liquidity providers.

\citet{goldstein_commodity_2017} as well as
\citet{goldstein_speculation_2014} argue that the arrival of financial
investors has an impact on the extent of normal backwardation in
commodity futures markets. Based on their analysis we would expect our
average CHP factor, which is based on backwardation measured by hedger's
positions, as well our term structure factor, which is based on
backwardation measured by roll yield, to be best suited to measure the
impact of financialization.\\
We analyze the pricing dynamics over the different periods by first
studying the time series pricing performance of the various factors on
all the US traded commodities with the factors constructed using the
latter set of commodities. The market based factor outperforms the other
three in the first period and it has a similar level of performance in
the second period, while the CHP factor shows the greatest improvement
in performance in the second period. The performance of the market
factor is driven by agricultural commodities and metals in both the
first and second period and the dramatic improvement of the CHP factor
in the second period is driven by precious and precious-base metals. The
best performing commodities for the market factor exhibited
approximately the same level of price co-movement both before and during
financialization. In contrast, the best performing commodities for the
CHP factor exhibited very low pair-wise correlation in the first period,
around the average for US traded commodities, while during
financialization the average pairwise correlation between the top
performing CHP factor commodities is higher than that for the market
factor and double of the average for all the US traded commodities. This
increase is due to the dramatic rise in average pairwise correlation
amongst the US traded metals all of which exhibited high
R\textsuperscript{2} with respect to the CHP factor during
financialization which in turn is due to their common low level of CHP
over this period. Thus the CHP factor model is able to establish the
strengthening link between a non-price attribute and price dynamics in
the US metal sub-sector during financialization.\\
We explore this issue further by constructing the four factors with the
top eight commodities for a given factor (factor picks) in each
period\footnote{See section \ref{methods} for further details.} and
evaluating its performance on these selected commodities as well as on
all US traded commodities. The results of this exercise strongly suggest
that the hedging pressure co-movement and price co-movement link was
concentrated in the metals sub-sector.

@@

Financialization was an issue of such policy importance that it
triggered legislative action. The debate was initially framed around
adequacy of speculation, the burning issue of the day, and the
consequent analysis focused on the more mechanical effects of
financialization. With the benefit of hindsight, it appears this
approach was perhaps too narrow and it now seems necessary to address
financialization from a broader perspective. Commodity price dynamics
appear to have altered substantially in quite different ways over the
financialization and the financial crisis periods and our commodity
futures based asset pricing approach seems able to provide new insights
into the nature of these changes: financialization was a phenomenon
endogenous to the commodity markets transmitted via the commodity
futures markets and mainly concentrated in a specific sector (metals)
while the crisis and its aftermath seems to have delivered an exogenous
shock across the entire cross-section of global commodities. The initial
view of financialization was that it consisted of speculative flows
which had the effect of driving up and creating bubble like conditions
in commodity future prices\footnote{\citet{de_schutter_food_2010_1};
  \citet{gilbert_how_2010}; \citet{gilbert_speculative_2010};
  \citet{masters_testimony_2008}; \citet{masters_accidental_2008};
  \citet{herman_not_2011_1}; \citet{schumann_hunger-makers_2011};
  \citet{singleton_investor_2013}; \citet{unctad_global_2009}.}.
However, the fundamental question of the nature of the impact of
financialization across the entire cross-section of commodity futures
markets has not yet been completely answered. We provide an empirical
complement to a new stream of theoretical studies that try to model the
impact of financialization on various aspects of commodity futures
markets\footnote{\citet{etula_broker-dealer_2013},
  \citet{acharya_limits_2013}, \citet{cheng_convective_2014},
  \citet{leclercq_equilibrium_2014}, \citet{sockin_informational_2015},
  \citet{goldstein_speculation_2013},
  \citet{EkelandSpeculationcommodityfutures2016},
  \citet{goldstein_commodity_2017}.} and demonstrate that the effects of
financialization extend beyond the mechanical effects induced by
indexation as outlined in \citet{basak_model_2016} as well as
\citet{tang_index_2012}.

@@

A world-wide debate has ensued about the role of index funds in
commodity markets. The first responses to the 2007/2008 crisis of
escalating food and energy prices took the form of policy reports, many
of which reasoned that the growth of commodity index funds came along
with an influx of largely speculative capital that was responsible for
driving commodity prices beyond their historic highs
\citep{de_schutter_food_2010_1, gilbert_speculative_2010, herman_not_2011_1, schumann_hunger-makers_2011, unctad_global_2009}.
The US and various european countries also became involved in this
issue. Early US Senate investigations drew on pricing and trading data
supplied by the CFTC as well as interviews with numerous experts
including hedge fund manager Michael Masters who linked the growing
presence of index investors in US wheat markets to the observed surges
in both futures and cash prices
\citep{masters_testimony_2008, masters_accidental_2008}). This
contention has since come to be known as the Masters' hypothesis and was
largely endorsed by the final US Senate report on the issue
\citep{USSenate_Excessive_speculation_wheat_2009}. At this stage, the
question thrust on the academic sphere was whether ``excessive
speculation'' (understood as the market activity peculiar to index type
investors) was linked to escalating energy and agricultural commodity
prices. A number of academic studies, most of which used commodity
specific causality and correlation based analysis have disproved this
contention, particularly in the case of agricultural and energy
commodities. \citet{irwin_devil_2009}, \citet{sanders_adequacy_2010},
\citet{sanders_impact_2011}, \citet{sanders_new_2011},
\citet{irwin_commodity_2013}, \citet{brunetti_commodity_2014},
\citet{hamilton_effects_2015} and \citet{bruno_financialization_2017}
study the agricultural markets. \citet{buyuksahin_speculators_2011},
\citet{tokic_speculation_2012}, \citet{fattouh_role_2013},
\citet{kilian_role_2014}, \citet{knittel_simple_2016} and
\citet{manera_modelling_2016} study the energy markets.
\citet{bohl_does_2013_2}, \citet{kim_does_2015} and
\citet{boyd_prevalence_2016} study both energy and agricultural markets
while \citet{irwin_index_2011}, \citet{irwin_financialization_2012} and
\citet{stoll_commodity_2011} study the commodity markets in general.

Other studies have underlined various alterations in commodity pricing
dynamics over the period. \citet{gilbert_role_2014} reject the view that
financialization has not had any effects in the grains markets and
demonstrate that trades originated by financial actors, and specifically
index investors, can move prices but tend typically to be
volatility-reducing. \citet{cheng_financialization_2014} show that in
the case of index commodities, financialization has transformed the risk
sharing, and information discovery functions of commodity futures
markets. \citet{juvenal_speculation_2015} contend that the oil price
increase between 2004 and 2008 was mainly driven by the strength of
global demand but that the financialization process of commodity markets
also played a role. \citet{henderson_new_2015} show that
non-information-based financial investments have important impacts on
commodity prices.

@@

Commodity index funds (CIFs), commodity based exchange traded funds
(ETFs) and notes (ETNs) as well as commodity linked notes (CLNs) were
amongst the main investment vehicules that allowed institutional and
retail investors to build up commodity exposure in the early 2000s
\citep{boons_stock_2012, henderson_new_2015, irwin_index_2011, unctad_global_2009, schumann_hunger-makers_2011}.
Most of these products were primarily designed to track prominent
commodity indexes amongst which the S\&P-GSCI, the most popular. Most of
its constituents at the time of writting form the basis of the broad
cross-section of commodities that we consider in this study\footnote{See
  section \ref{assets} for details on the commodity pool considered.}.
Unleaded gasoline, the NYMEX legacy contract for gasoline stopped
trading in 2006 when it was replaced by Reformulated Gasoline Blendstock
for Oxygen Blending (RBOB) with the two futures series trading alongside
for most of the year. For our NYMEX gasoline data series, we consider
unleaded gasoline up to September 1\textsuperscript{st}, 2006 and RBOB
gasoline thereafter, date at which liquidity for the latter overtook
that for the former.

>>>>

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Financialization of Commodities: an Asset Pricing Perspective},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Financialization of Commodities: an Asset Pricing Perspective}
    \author[strathclyde]{Devraj Basu}
   \ead{devraj.basu@strath.ac.uk}

    \author[strathclyde]{Olivier Bauthéac}
   \ead{olivier.bautheac@strath.ac.uk}

      \address[strathclyde]{University of Strathclyde, Glasgow, UK}

  \begin{abstract}
  The last two decades have seen major changes in the nature of the
  commodity markets with the influx of financial investor inflows followed
  by the financial crisis. We study the effect of these events on
  commodity futures pricing dynamics using commodity asset pricing
  techniques. We find that a non-price based factor is best able to detect
  the effects of commodity financialization, while the effects of the
  financial crisis and after are most clearly visible through the
  performance of the market factor. The effects of financialization seem
  to go beyond mechanical effects induced by the rise of indexing.
  \end{abstract}
   \begin{keyword} Commodity futures, financialization, hedging pressure, asset pricing\end{keyword}
 \end{frontmatter}

@@

\begin{itemize}
\tightlist
\item
  Devraj Basu: I am a co-author of this paper and wish to state that I
  have no conflict of interest, financial or otherwise. I do not hold
  paid or unpaid positions at any organization whose goals relate to the
  article.\\
  \includegraphics{Devraj.jpg}
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Olivier Bauthéac: I am a co-author of this paper and wish to state
  that I have no conflict of interest, financial or otherwise. I do not
  hold paid or unpaid positions at any organization whose goals relate
  to the article.\\
  \includegraphics{Olivier.jpg}
\end{itemize}

@@

The last twenty years have seen major upheavals in the commodity markets
with the onset of financialization\footnote{Starting around 2004, with a
  view of commodity markets as an emerging asset class and in an effort
  to hedge against inflation and seek diversification benefits
  (Büyükşahin and Robe, 2014; Singleton, 2013), institutional investors
  sent forth an unprecedented flow of capital into commodity futures
  markets. This phenomenon is commonly referred to as the
  financialization of commodities (Domanski and Heath, 2007).} as well
as the period of the 2008 financial crisis and its aftermath.
Financialization appears to have affected a number of economic sectors,
including agriculture and energy, and its impact has been hotly debated
in the policy, legislative, regulatory, and academic spheres. A vigorous
legislative debate has led to regulatory changes accompanied by an
equally intense policy and academic debate across a number of
disciplines about the perceived effects on commodity prices of the
unprecedented inflow of institutional funds brought into commodity
futures markets by new index type investors\footnote{See section
  \ref{background} for a detailled discussion.}. In response to
immediate policy concerns, the academic debate was initially framed
rather narrowly. In this paper we take a fundamentally different
empirical approach that is broader in scope. We consider the entire
cross-section of actively traded commodities on futures markets and use
a futures based asset pricing framework that includes factors
constructed using both price and non-price attributes.

@@

The first factor model we consider (CHP) is based on the well-known
hedging pressure based theory (Anderson and Danthine, 1983; Chang, 1985;
Cootner, 1960; Dusak, 1973; Hicks, 1939; Hirshleifer, 1990, 1989, 1988;
Keynes, 1930; Kolb, 1992) which postulates that futures prices for a
given commodity are inversely related to the extent that commercial
hedgers are short or long and we use a factor mimicking portfolio that
captures the impact of hedging pressure as a systemic factor (Basu and
Miffre, 2013). Our second factor model is based on open interest growth
following Hong and Yogo (2012) who find that open interest growth has
predictive power for individual commodity returns and again we construct
a long-short factor mimicking portfolio. Our third model is based on
term structure following Szymanowska et al. (2014) and Fuertes et al.
(2015) who demonstrate that the term structure factor has explanatory
power for the cross-section of commodity returns. We finally consider an
equally weighted portfolio of commodities as a proxy for the market
factor. Our two price based and two non-price based factors allow us to
compare and contrast different channels that could have led to changes
in commodity price dynamics over the last twenty years.

We study the performance of these models on a set of twenty-four US
traded commodities and six UK traded (LME exchange) base metals over the
1997/2017 period that we divide into four distinct sub-periods. The
first, the pre-financialization phase, spans the 1997/2003 period; the
second, the financialization phase, spans the 2004/2007 period\footnote{Starting
  point based on earlier studies (Baker, 2014; Christoffersen et al.,
  2014).}; the third, 2008/2013, encompasses both the financial crisis
and the loose monetary policy regimes that followed; and the fourth
spans the period starting in 2014 to present.\\
Consistent with earlier studies, we observe that mean returns for the US
traded commodities are sharply higher over the financialization period
compared to pre-financialization and this pattern extends to UK traded
metals.

The cross-sectional nature of financialization (Basak and Pavlova, 2016;
Cheng and Xiong, 2014) leads us to extend the Keynesian individual
hedging pressure paradigm by incorporating aggregate CHP (\textbf{CHP}),
a market wide measure of hedging pressure first considered in Hong and
Yogo (2012). This paradigm implies that returns for individual
commodities should be high in periods where \textbf{CHP} is low
(\textbf{backwardation}) and low in periods where \textbf{CHP} is high
(\textbf{contango}). Over the first period the cross sectional average
returns during \textbf{backwardation} for both the US and UK commodities
are positive with those for the US statistically significant, and
significantly higher than during \textbf{contango}, providing broad
support for an aggregate version of the Keynesian hedging pressure
hypothesis. Over the second period this pattern reverses, with both the
US and UK commodities having significant positive returns during
\textbf{contango} with these returns being significantly higher than
during **backwardation\}. This change in pattern suggests that the onset
of financialization altered the traditional risk shifting nature of
hedger's behaviour, which is implied by the Keynesian hypothesis. The
arrival of financial investors raises the question of whether they
functioned as liquidity providers or liquidity demanders. Arguments have
been made for both of these possibilities with Moskowitz et al. (2012)
arguing that financial investors provide liquidity while Kang et al.
(2017) argue that they demand liquidity. Our analysis of returns in
periods of low and high aggregate CHP before and during financialization
allows us to analyse this issue from a new perspective and suggests that
financial investors may have been liquidity demanders leading hedgers to
become liquidity providers.

Goldstein and Yang (2017) as well as Goldstein et al. (2014) argue that
the arrival of financial investors has an impact on the extent of normal
backwardation in commodity futures markets. Based on their analysis we
would expect our average CHP factor, which is based on backwardation
measured by hedger's positions, as well our term structure factor, which
is based on backwardation measured by roll yield, to be best suited to
measure the impact of financialization.\\
We analyze the pricing dynamics over the different periods by first
studying the time series pricing performance of the various factors on
all the US traded commodities with the factors constructed using the
latter set of commodities. The market based factor outperforms the other
three in the first period and it has a similar level of performance in
the second period, while the CHP factor shows the greatest improvement
in performance in the second period. The performance of the market
factor is driven by agricultural commodities and metals in both the
first and second period and the dramatic improvement of the CHP factor
in the second period is driven by precious and precious-base metals. The
best performing commodities for the market factor exhibited
approximately the same level of price co-movement both before and during
financialization. In contrast, the best performing commodities for the
CHP factor exhibited very low pair-wise correlation in the first period,
around the average for US traded commodities, while during
financialization the average pairwise correlation between the top
performing CHP factor commodities is higher than that for the market
factor and double of the average for all the US traded commodities. This
increase is due to the dramatic rise in average pairwise correlation
amongst the US traded metals all of which exhibited high
R\textsuperscript{2} with respect to the CHP factor during
financialization which in turn is due to their common low level of CHP
over this period. Thus the CHP factor model is able to establish the
strengthening link between a non-price attribute and price dynamics in
the US metal sub-sector during financialization.\\
We explore this issue further by constructing the four factors with the
top eight commodities for a given factor (factor picks) in each
period\footnote{See section \ref{methods} for further details.} and
evaluating its performance on these selected commodities as well as on
all US traded commodities. The results of this exercise strongly suggest
that the hedging pressure co-movement and price co-movement link was
concentrated in the metals sub-sector.

@@

Financialization was an issue of such policy importance that it
triggered legislative action. The debate was initially framed around
adequacy of speculation, the burning issue of the day, and the
consequent analysis focused on the more mechanical effects of
financialization. With the benefit of hindsight, it appears this
approach was perhaps too narrow and it now seems necessary to address
financialization from a broader perspective. Commodity price dynamics
appear to have altered substantially in quite different ways over the
financialization and the financial crisis periods and our commodity
futures based asset pricing approach seems able to provide new insights
into the nature of these changes: financialization was a phenomenon
endogenous to the commodity markets transmitted via the commodity
futures markets and mainly concentrated in a specific sector (metals)
while the crisis and its aftermath seems to have delivered an exogenous
shock across the entire cross-section of global commodities. The initial
view of financialization was that it consisted of speculative flows
which had the effect of driving up and creating bubble like conditions
in commodity future prices\footnote{De Schutter (2010); Gilbert (2010a);
  Gilbert (2010b); Masters (2008); Masters and White (2008); Herman et
  al. (2011); Schumann (2011); Singleton (2013); UNCTAD and Cooperation
  (2009).}. However, the fundamental question of the nature of the
impact of financialization across the entire cross-section of commodity
futures markets has not yet been completely answered. We provide an
empirical complement to a new stream of theoretical studies that try to
model the impact of financialization on various aspects of commodity
futures markets\footnote{Etula (2013), Acharya et al. (2013), Cheng et
  al. (2014), Leclercq and Praz (2014), Sockin and Xiong (2015),
  Goldstein et al. (2014), Ekeland et al. (2016), Goldstein and Yang
  (2017).} and demonstrate that the effects of financialization extend
beyond the mechanical effects induced by indexation as outlined in Basak
and Pavlova (2016) as well as Tang and Xiong (2012).

@@

A world-wide debate has ensued about the role of index funds in
commodity markets. The first responses to the 2007/2008 crisis of
escalating food and energy prices took the form of policy reports, many
of which reasoned that the growth of commodity index funds came along
with an influx of largely speculative capital that was responsible for
driving commodity prices beyond their historic highs (De Schutter, 2010;
Gilbert, 2010b; Herman et al., 2011; Schumann, 2011; UNCTAD and
Cooperation, 2009). The US and various european countries also became
involved in this issue. Early US Senate investigations drew on pricing
and trading data supplied by the CFTC as well as interviews with
numerous experts including hedge fund manager Michael Masters who linked
the growing presence of index investors in US wheat markets to the
observed surges in both futures and cash prices (Masters, 2008; Masters
and White, 2008)). This contention has since come to be known as the
Masters' hypothesis and was largely endorsed by the final US Senate
report on the issue (Senate, 2009). At this stage, the question thrust
on the academic sphere was whether ``excessive speculation'' (understood
as the market activity peculiar to index type investors) was linked to
escalating energy and agricultural commodity prices. A number of
academic studies, most of which used commodity specific causality and
correlation based analysis have disproved this contention, particularly
in the case of agricultural and energy commodities. Irwin et al. (2009),
Sanders et al. (2010), Sanders and Irwin (2011a), Sanders and Irwin
(2011b), Irwin (2013), Brunetti and Reiffen (2014), Hamilton and Wu
(2015) and Bruno et al. (2017) study the agricultural markets.
Büyükşahin and Harris (2011), Tokic (2012), Fattouh et al. (2013),
Kilian and Murphy (2014), Knittel and Pindyck (2016) and Manera et al.
(2016) study the energy markets. Bohl and Stephan (2013), Kim (2015) and
Boyd et al. (2016) study both energy and agricultural markets while
Irwin and Sanders (2011), Irwin and Sanders (2012) and Stoll and Whaley
(2011) study the commodity markets in general.

Other studies have underlined various alterations in commodity pricing
dynamics over the period. Gilbert and Pfuderer (2014) reject the view
that financialization has not had any effects in the grains markets and
demonstrate that trades originated by financial actors, and specifically
index investors, can move prices but tend typically to be
volatility-reducing. Cheng and Xiong (2014) show that in the case of
index commodities, financialization has transformed the risk sharing,
and information discovery functions of commodity futures markets.
Juvenal and Petrella (2015) contend that the oil price increase between
2004 and 2008 was mainly driven by the strength of global demand but
that the financialization process of commodity markets also played a
role. Henderson et al. (2015) show that non-information-based financial
investments have important impacts on commodity prices.

@@

Commodity index funds (CIFs), commodity based exchange traded funds
(ETFs) and notes (ETNs) as well as commodity linked notes (CLNs) were
amongst the main investment vehicules that allowed institutional and
retail investors to build up commodity exposure in the early 2000s
(Boons et al., 2012; Henderson et al., 2015; Irwin and Sanders, 2011;
Schumann, 2011; UNCTAD and Cooperation, 2009). Most of these products
were primarily designed to track prominent commodity indexes amongst
which the S\&P-GSCI, the most popular. Most of its constituents at the
time of writting form the basis of the broad cross-section of
commodities that we consider in this study\footnote{See section
  \ref{assets} for details on the commodity pool considered.}. Unleaded
gasoline, the NYMEX legacy contract for gasoline stopped trading in 2006
when it was replaced by Reformulated Gasoline Blendstock for Oxygen
Blending (RBOB) with the two futures series trading alongside for most
of the year. For our NYMEX gasoline data series, we consider unleaded
gasoline up to September 1\textsuperscript{st}, 2006 and RBOB gasoline
thereafter, date at which liquidity for the latter overtook that for the
former.
